Natural Language Processing\\ Fall 2015 Final Project:\\ Sense Disambiguation with the\\ Na\"ive Bayes Classifier

### 2015-12-15 paper/paper.tex
Remove erroneous graph inclusion

\begin{abstract}
This paper examines an attempt to implement the Na\"ive Bayes Classifier algorithm for the purpose of performing the task of Word Sense Disambiguation. In it we will discuss the problem of Word Sense Disambiguation broadly, along with a thorough explanation of the Na\"ive Bayes algorithm. Finally, we will present an implementation of the algorithm in Python and analyze the results of the application of the alogrithm and difficulties that arose in its application.
\end{abstract}

@@

The second form is derived using Bayes' Formula, hence Na\"ive \emph{Bayes} Classifier. This equation is simple and conceptually effective. But in practice, even the second form is difficult to use---meainingfully computing the probability that a given feature vector will occur requires that that entire feature vector have been encountered for that sense before. For large feature vectors the likelihood of this is extremely low, and requires prohibitively large sets of training data. The Na\"ive Bayes Classifier (NBC) is an attempt to minimize the combinations of the features by making the \emph{na\"ive} assumption that the occurrence of each feature is an independent event. With this assumption, we may say that ${P(F(w)|s) = \prod_{f \in F(w)} P(f|s)}$. This assumption is certainly not always true, and is what gives the NBC its na\"ive qualifier. Indeed this equation is often an approximation at best, but usually it is sufficiently accurate to be useable.

@@

\subsection{Baseline}
An important part of establishing the performance of an algorithm is to determine the most basic implmentation that the algorithm should improve upon. We call this the baseline. In our case, we chose to implement a most-frequent-sense selector as our baseline upon which to improve. This algorithm simply trains on a set of sense-annotated data and determines which sense is most common for each word. When disambiguating, the algorithm simply chooses that most-frequent sense for each word.

@@

For meaningful feature extraction, it is helpful to attempt to obtain additional metadata about the input that will be sense-disambiguated. This is called the pre-processing stage, the goal of which to attempt to determine as much data that will be useful for identifying sense-context as possible. To this effect, we utilized the Natural Language ToolKit for lemmatization and part-of-speech tagging. These two basic pieces of information can immediately rule out most plainly incorrect senses from being chosen.

@@

\section{Analysis}
Overall, the accuracy of the Na\"ive Bayes Classifier in our implementation was disappointing and our results were somehwat inconsistent with what we expected. We were unable to identify any features that could be calculated within the scope of the project that significantly improved the accuracy of the sense disambiguation.

@@

<LINE GRAPH OF MISSED PERCENTAGE PER FILE COMPARED WITH BASELINE>

<LINE GRAPH OF INCORRECT PERCENTAGE PER FILE COMPARED WITH BASELINE>

<BAR GRAPH OF AVERAGE CORRECT, MISSED, INCORRECT, COMPARED WITH BASELINE>

@@

\section{Conclusion}
The accuracy of the results obtained from the system were signficantly less than we hoped. The difficulty in identifying and successfully extracting features that actually provided meaningful data that improved results was unexpected. This difficulty stemmed from the fact that the Na\"ive Bayes Classifier is difficult to reason about, as a probabilistic method. Choices are made as a unit, the product of a large number of factors that are calculated from large amounts of data. This is why other methods, like Decision Trees, are often preferred when developing Word Sense Disambiguation systems; it is generally easier to understand and follow the choices that those systems make, making tuning easier.

>>>>

\begin{abstract}
This paper examines an attempt to implement the Na\"ive Bayes Classifier algorithm for the purpose of performing the task of Word Sense Disambiguation. In it we will discuss the problem of Word Sense Disambiguation broadly, along with a thorough explanation of the Na\"ive Bayes algorithm. Finally, we will present an implementation of the algorithm in Python and analyze the results of the application of the algorithm and difficulties that arose in its application.
\end{abstract}

@@

The second form is derived using Bayes' Formula, hence Na\"ive \emph{Bayes} Classifier. This equation is simple and conceptually effective. But in practice, even the second form is difficult to use---meaningfully computing the probability that a given feature vector will occur requires that that entire feature vector have been encountered for that sense before. For large feature vectors the likelihood of this is extremely low, and requires prohibitively large sets of training data. The Na\"ive Bayes Classifier (NBC) is an attempt to minimize the combinations of the features by making the \emph{na\"ive} assumption that the occurrence of each feature is an independent event. With this assumption, we may say that ${P(F(w)|s) = \prod_{f \in F(w)} P(f|s)}$. This assumption is certainly not always true, and is what gives the NBC its na\"ive qualifier. Indeed this equation is often an approximation at best, but usually it is sufficiently accurate to be usable.

@@

\subsection{Baseline}
An important part of establishing the performance of an algorithm is to determine the most basic implementation that the algorithm should improve upon. We call this the baseline. In our case, we chose to implement a most-frequent-sense selector as our baseline upon which to improve. This algorithm simply trains on a set of sense-annotated data and determines which sense is most common for each word. When disambiguating, the algorithm simply chooses that most-frequent sense for each word.

@@

For meaningful feature extraction, it is helpful to attempt to obtain additional metadata about the input that will be sense-disambiguated. This is called the pre-processing stage, the goal of which to attempt to determine as much data that will be useful for identifying sense-context as possible. To this effect, we utilized the Natural Language Toolkit for lemmatization and part-of-speech tagging. These two basic pieces of information can immediately rule out most plainly incorrect senses from being chosen.

@@

\section{Analysis}
Overall, the accuracy of the Na\"ive Bayes Classifier in our implementation was disappointing and our results were somewhat inconsistent with what we expected. We were unable to identify any features that could be calculated within the scope of the project that significantly improved the accuracy of the sense disambiguation.

@@

@@

\section{Conclusion}
The accuracy of the results obtained from the system were significantly less than we hoped. The difficulty in identifying and successfully extracting features that actually provided meaningful data that improved results was unexpected. This difficulty stemmed from the fact that the Na\"ive Bayes Classifier is difficult to reason about, as a probabilistic method. Choices are made as a unit, the product of a large number of factors that are calculated from large amounts of data. This is why other methods, like Decision Trees, are often preferred when developing Word Sense Disambiguation systems; it is generally easier to understand and follow the choices that those systems make, making tuning easier.
